---
title: "Apartment Rental Prediction System"
author: "Dr. Ivan S. Zapreev"
date: "`r format(Sys.Date())`"
output: pdf_document
---

```{r setup, include=FALSE}
if(!require(knitr)) install.packages("knitr", repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra", repos = "http://cran.us.r-project.org")
if(!require(graphics)) install.packages("graphics", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(tidyr)) install.packages("tidyr", repos = "http://cran.us.r-project.org")

knitr::opts_chunk$set(echo = TRUE)

library(tidyr)
library(tidyverse)
library(lubridate)
library(graphics)
library(kableExtra)
library(dslabs)

options(digits=7)

MOVIELENS_DATA_FILE_NAME <- "arog_data.rda"
MOVIELENS_REPORT_FILE_NAME <- "arog_report.rda"

#Load the files
load(MOVIELENS_DATA_FILE_NAME)
load(MOVIELENS_REPORT_FILE_NAME)

#--------------------------------------------------------------------
# This function counts the number of N/A entries in the columns of 
# the provided data set, it returns a tibble with the columns:
#    name - storing the column name of the data set
#    count - storing the number of N/A entries
#    percent - storing the percent of N/A entries
#--------------------------------------------------------------------
count_na_entries <- function(data_set) {
  col_names <- names(data_set)
  na_counts <- as.vector(sapply(col_names, function(name){
    return(sum(is.na(data_set[,name])))
  }))
  num_rows <- nrow(data_set)
  percents <- round(na_counts*100/num_rows, 2)
  result <- tibble(name = col_names, 
                count = na_counts,
                percent = percents) %>%
    arrange(desc(count))
  names(result) <- c("Column name", "N/A count", "N/A percent")
  return(result)
}
```

# Introduction
<!-- An introduction/overview/executive summary section that:
 1. Describes the dataset
 2. Summarizes the goal of the project 
 3. Outlines the key steps that were performed; -->


## Dataset overview

As stated on the [webpage](`r arog_report$AROG_DATA_SET_SITE_URL`) of the `r arog_report$AROG_DATA_SET_NAME`, it contains `198,379` rental offers scraped from the Germany's biggest real estate online platform ß [ImmobilienScout24](https://www.immobilienscout24.de/). 

The data set consists of a single CSV file: *`r arog_report$AROG_CSV_FILE_NAME`* which only contains offers for rental properties. The data features important rental property attributes, such as the living area size, the rent (both base rent as well as total rent), the location, type of energy, and etc. The `date` column present in the data set defines the time of scraping, which was done on three distinct dates: *2018-09-22*, *2019-05-10* and *2019-10-08*.

The complete list of data set columns is extensive[^1] and thus in this study we will use the following subset:
```{r, echo=F}
names(arog_data$selected_data)
```
This sub-selection reduces the number of considered data set columns[^2] from $48$ to $26$ and is motivated by the personal preferences of the report's author and has no scientifically proven motivation. On the contrary, this column selection shall be seen as a part of problem statement. In other words, the task is to build an accurate[^3] rental price prediction model based on the predictors from this set of columns.

The additional data preparation steps will be described in the *"Data wrangling"* section of this document.

[^1]: Please consider reading *"Appendix A"* for the complete list of the data set columns.
[^2]: Please consider reading *"Appendix B"* for the column descriptions.
[^3]: Please consider reading the *"Project goal"* section for an exact goal formulation.

## Project goal



## Execution plan


# Data wrangling
<!-- A data wrangling section that explains how the data was prepared for data analysis -->
In this section we present cleaning, restructuring and enriching the raw data taken from the `r arog_report$AROG_DATA_SET_NAME`.

First, let us note that the number of data entries in the original data set is equal to `r nrow(arog_data$selected_data)`. This data is however not ready to be worked with as it contains multiple `N/A` values and other inconsistencies. For example, the next table summarizes the number of `N/A` values per column:

```{r, echo=F}
sel_data_na_cnts <- count_na_entries(arog_data$selected_data)

sel_data_na_cnts%>% print(., n=Inf)
```
As one can see, about $\frac{1}{2}$ of the columns has $10$--$80$% `N/A`$^{s}$, whereas the other half has (almost) no `N/A`$^{s}$.

The remainder of the section will be organized as follows. First we explain how we cleaned the data and solved some of its inconsistencies. Then we provide a summary of the cleaned data set. In the end we explain how we split the entire data set into the `validation` and `modeling` sub-sets[^4].

[^4]:The latter will also be split into the `training` and `testing` set for the sake of model cross-validation.

## Data cleaning
The data cleaning will be explained in the next steps:

1. We begin with the `totalRent` column as this is the value that we want to predict;
2. We proceed with the columns with the marginal ($< 1$%) of `N/A` values; 
3. We cover the remaining columns in the descending order of the number of `N/A` values.
4. We consider and sole some other data inconsistencies.

```{r, echo = F}
idx <- which(sel_data_na_cnts[,"Column name"] == "totalRent")
total_rent_na_percent <- sel_data_na_cnts[idx, "N/A percent"]
num_marginal_columns <- nrow(filter(sel_data_na_cnts, sel_data_na_cnts[,"N/A percent"] < 1))
```

The `totalRent` column contains data that we want to predict. Therefore, the rows with `totalRent` == `N/A` are useless to us and shall be removed. Unfortunately, this will reduce the data set by `r total_rent_na_percent`%. There are also `r num_marginal_columns` columns with a marginal ($0$ to $1$) number of `N/A` values. The latter can be seamlessly removed as even if all of these `N/A`$^{s}$ appear in different rows, we will remove at most `r num_marginal_columns` entries which is `r round(num_marginal_columns/nrow(arog_data$selected_data)*100,4)`% of data.

### The remaining columns



### Additional issues

## Clean data summary
Let us now summarize the resulting clean data:

```{r, echo=F}
clean_data_na_cnts <- count_na_entries(arog_data$cleaned_data)
clean_data_na_cnts%>% print(., n=Inf)
```

```{r, echo = F}
na_total_rent_cnt <- round(nrow(arog_data$selected_data)/100*total_rent_na_percent)
selected_wtr <- nrow(arog_data$selected_data) - na_total_rent_cnt
```

As one can notice, the dat set size has been reduced from `r nrow(arog_data$cleaned_data)` to `r  nrow(arog_data$selected_data)`. The major reason for that is excluding the rows with the `N/A` values of the `totalRent` column. Let us recall that the number of such raws was `r total_rent_na_percent`% of the data set, e.g. `r na_total_rent_cnt` rows. It now remains to notice that `r nrow(arog_data$selected_data)` $-$ `r na_total_rent_cnt` $=$ `r selected_wtr` $\approx$ `r nrow(arog_data$cleaned_data)`. The remaining delta is explained by cleaning the `floor`/`typeOfFlat` column inconsistencies.

## Splitting data

# Data analysis
<!-- A data analysis section that:
 1. Explains the process and techniques used, such as:
  1.1 Data exploration
  1.2 Data visualization
 2. Presents insights gained -->

# Modeling approach
<!-- A modeling approach section that explains the modeling approach -->



# Results
<!-- A results section that:
 1. Presents the modeling results 
 2. Discusses the model performance -->



# Conclusions
<!-- A conclusion section that:
 1. Gives a brief summary of the report
 2. Explains its limitations
 3. Suggests future work -->

# Appendix A: The complete list of data set columns

Hereby we present the list of columns from the original data set:

```{r, echo = F}
names(arog_data$raw_data)
```

# Appendix B: Data set column descriptions

Here is the list of the initially considered data set columns with the descriptions thereof:

1. `hasKitchen` -- has a kitchen
2. `balcony` -- does the object have a balcony
3. `cellar` -- has a cellar
4. `lift` -- is elevator available
5. `floor` -- which floor is the flat on
6. `garden` -- has a garden
7. `noParkSpaces` -- number of parking spaces
8. `livingSpace` -- living space in sqm
9. `condition` -- condition of the flat
10. `interiorQual` -- interior quality

11. `regio1` --  Bundesland
12. `regio2` - District or Kreis, same as geo krs
13. `regio3` -- City/town

14. `noRooms` -- number of rooms
15. `numberOfFloors` -- number of floors in the building
16. `typeOfFlat` -- type of flat
17. `yearConstructed` -- construction year
18. `newlyConst` -- is the building newly constructed
19. `heatingType` -- Type of heating
20. `energyEfficiencyClass` -- energy efficiency class

21. `heatingCosts` -- monthly heating costs in €
22. `serviceCharge` --  auxiliary costs such as electricity or Internet in €
23. `electricityBasePrice` -- monthly base price for electricity in €
24. `baseRent` -- base rent without electricity and heating
25. `totalRent` -- total rent (usually a sum of base rent, service charge and heating cost)

26. `date` -- time of scraping
